NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids

### 2018-02-19 Manuscript/manuscriptPS.tex
Update manuscript

%\label{}
\section{Introduction}
Phosphatidylserine (PS) is the most common negatively
charged lipid in eykaryotic membranes.
PS lipids compose 8.5\% of total lipid weight of erythrocytes,
but the abundance varies between different organelles up to
25-35\% in plasma membrane \cite{lemmon08,leventis10,li14}.
Despite of the relatively low abundance, PS lipids
are important signaling molecules. They interact with
signaling proteins \cite{leventis10}, regulate
surface charge and protein localization \cite{yeung08}, and
induce protein aggregation \cite{zhao04,??}.
Some domains spesifically interact PS lipids,
while others are attracted by general electrostatics and the
binding can be regulated by calcium \cite{leventis10}.
Therefore, it is crucial to resolve the structural details
of PS headgroup and calcium binding in order to fully
understand PS mediated signaling processes.

@@

PC lipid headgroup order parameters can used to measure ion binding
affinity, because their magnitude is proportional
to the amount of bound charge in bilayer \cite{seelig87,catte16}.
The molecular electrometer concept can be used also
for bilayers containing PC lipids mixed with charged lipids \cite{borle85,macdonald87,roux90}.
This is demonstrated in Fig. %s \ref{OrderParametersWithCaCl},
%\ref{OrderParametersWithCaClBELOW1M} and
\ref{OrderParameterCHANGESWithCaClBELOW1M},
showing the changes of PC headgroup order parameters %for PC headgroup $\alpha$ and $\beta$ carbons
as a function of CaCl$_2$ concentration in the presence of different amounts of
negatively charged PS or PG lipids.

%PC headgroup order parameters increase when negatively charged
%PS or PG are added to PC bilayer in the absense of added CaCl$_2$,
%as expected based on electrometer concept \cite{seelig87}
%(see Fig. \ref{OrderParametersWithCaClBELOW1M}).
%In electrometer concept this is explained by the tilting of
%headgroup more parallel to membrane normal \cite{??}.
The decrease of order parameters with CaCl$_2$
is more pronounced for systems with more
negatively charged lipids.% (see Fig. \ref{OrderParameterCHANGESWithCaClBELOW1M}).
Order parameters reach the values of pure PC bilayer close to CaCl$_2$ concentrations of $\sim$ 50-300mM.
At this point the Ca2+ binding presumably fully cancels the charge from negative lipids and
overcharging occurs above these concenterations.
The interpretation of this data and some other results has been that \cite{seelig90}
\begin{displayquote}
  {\it ''(i) Ca$^{2+}$ binds to neutral lipids (phosphatidylcholine, phosphatidylethanolamine) and negatively charged lipids
    (phosphatidylglycerol) with approximately the same binding constant of K = 10-20 M$^{-1}$; \\
    (ii) the free Ca$^{2+}$
    concentration at the membrane interface is distinctly enhanced if the membrane carries a negative surface
    charge, either due to protein or to lipid; \\
    (iii) increased inter-facial Ca$^{2+}$ also means increased amounts
    of bound Ca$^{2+}$ at neutral and charged lipids; \\
    (iv) the actual binding step can be described by a Langmuir
    adsorption isotherm with a 1 lipid:1 Ca$^{2+}$ stoichiometry, provided the interfacial concentration C$_M$, is
    used to describe the chemical binding equilibrium.''}
\end{displayquote}

@@

Before using headgroup order parameters to compare ion binding affinity between simulations
and experiments, the response of the order parameters to bound charge has to be quantified.
The response of headgroup order parameters to the fixed amount of cationic surfactants in
POPC bilayer is compared between simulations and experiments \cite{scherer89} In Fig. \ref{CHANGESwithCaClPGPS}.
The figure shows that the order parameters are too sensitive to bound charge in Lipid14 model,
while CHARMM36 is in better agreement with experiments. This has to be taken into account when
analysin the binding affinities.
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGopsDHMDMAB.eps}
  \caption{\label{CHANGESwithCaClPGPS}
  The response of headgroup order parameters to the fixed amount of cationic surfactants in
  POPC bilayer is compared between simulations and experiments \cite{scherer89}.}
\end{figure}

>>>>

%\label{}
\section{Introduction}
Phosphatidylserine (PS) is the most common negatively
charged lipid in eykaryotic membranes.
PS lipids compose 8.5\% of total lipid weight of erythrocytes,
but the abundance varies between different organelles up to
25-35\% in plasma membrane \cite{lemmon08,leventis10,li14}.
Despite of the relatively low abundance, PS lipids
are important signaling molecules. They interact with
signaling proteins \cite{leventis10}, regulate
surface charge and protein localization \cite{yeung08}, and
induce protein aggregation \cite{zhao04,??}.
Some domains spesifically interact PS lipids,
while others are attracted by general electrostatics and the
binding can be regulated by calcium \cite{leventis10}.
Therefore, the structural details
of lipid headgroups and the details of calcium binding
are crucial for the PS mediated signaling processes.

@@

%The interpretation of this data and some other results has been that \cite{seelig90}
%\begin{displayquote}
%  {\it ''(i) Ca$^{2+}$ binds to neutral lipids (phosphatidylcholine, phosphatidylethanolamine) and negatively charged lipids
%    (phosphatidylglycerol) with approximately the same binding constant of K = 10-20 M$^{-1}$; \\
%    (ii) the free Ca$^{2+}$
%    concentration at the membrane interface is distinctly enhanced if the membrane carries a negative surface
%    charge, either due to protein or to lipid; \\
%    (iii) increased inter-facial Ca$^{2+}$ also means increased amounts
%    of bound Ca$^{2+}$ at neutral and charged lipids; \\
%    (iv) the actual binding step can be described by a Langmuir
%    adsorption isotherm with a 1 lipid:1 Ca$^{2+}$ stoichiometry, provided the interfacial concentration C$_M$, is
%    used to describe the chemical binding equilibrium.''}
%\end{displayquote}

@@

The headgroup order parameters of PC lipids can be used to measure ion binding
affinity because they change proportionally
to the amount of bound charge in bilayer \cite{seelig87,catte16}.
This molecular electrometer concept can be applied also to
lipid bilayers with mixtures of PC and negatively charged
lipids \cite{borle85,macdonald87,roux90}.
This is demonstrated in Fig. %s \ref{OrderParametersWithCaCl},
%\ref{OrderParametersWithCaClBELOW1M} and
\ref{OrderParameterCHANGESWithCaClBELOW1M}
using the previuosly reported experimental data
for mixtures with varying proportions of
negatively charged PS or PG lipids.
%
%the changes of PC headgroup order parameters %for PC headgroup $\alpha$ and $\beta$ carbons
%as a function of CaCl$_2$ concentration are shown
%
As expected, the decrease of order parameters with CaCl$_2$
is more pronounced for systems with more
negatively charged lipids.
% (see Fig. \ref{OrderParameterCHANGESWithCaClBELOW1M}).

%PC headgroup order parameters increase when negatively charged
%PS or PG are added to PC bilayer in the absense of added CaCl$_2$,
%as expected based on electrometer concept \cite{seelig87}
%(see Fig. \ref{OrderParametersWithCaClBELOW1M}).
%In electrometer concept this is explained by the tilting of
%headgroup more parallel to membrane normal \cite{??}.

%Order parameters reach the values of pure PC bilayer close to CaCl$_2$ concentrations of $\sim$ 50-300mM.
%At this point the Ca2+ binding presumably fully cancels the charge from negative lipids and
%overcharging occurs above these concenterations.

Before using the headgroup order parameters to compare ion binding affinity between simulations
and experiments, it is important to quantify the response of the order parameters to the
bound charge in simulations.
The response of headgroup order parameters to the fixed amount of cationic surfactants in
POPC bilayer is compared between simulations and experiments \cite{scherer89} In Fig. \ref{CHANGESwithCaClPGPS}.
The figure shows that the order parameters are too sensitive to bound charge in Lipid14 model,
while CHARMM36 is in better agreement with experiments. This has to be taken into account when
analysin the binding affinities.
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGopsDHMDMAB.eps}
  \caption{\label{CHANGESwithCaClPGPS}
  The response of headgroup order parameters to the fixed amount of cationic surfactants in
  POPC bilayer is compared between simulations and experiments \cite{scherer89}.}
\end{figure}
